Normalization by Evaluation for Sized Dependent Types

### 2017-06-10 nbe/icfp17.tex
[ ICFP 17 ] Some polishing in Section 4.

In this section we present an operational semantics of our language,
define the NbE algorithm, construct a PER model, and demonstrate that
NbE is complete for definitional equality, \ie, if
$\Gamma \der t = t' : T$, then $t$ and $t'$ have the same normal form.
\[
\begin{array}{lllll@{\qquad}l}
  \Ne & \ni &
  m & ::= & \ind i \mid m\,v \mid m\,a \mid m\,\ann a \mid
            m\,\case[\ell] V{v_z}{v_s} \mid m\,\fix[\ell] V v
    & \mbox{normal neutral term} \\
  \Nf & \ni &
  v & ::= & m \mid \lambda v \mid \zero a \mid \suc a v \mid
            \Set[\ell] \mid \Nat a \mid \piT {V_u}{V_t} \mid
            \epiT\Size V
    & \mbox{normal form} \\
\end{array}
\]
For the operational semantics,
instead of defining a separate language of values,
we extend the syntax of expressions by de Bruijn levels $\lev k$
to be used as generic values (unknowns), and type annotations
$\up A n$ and $\down A t$ for lazy realizations of the
reflection and reification operations of NbE.
\emph{Terms} are expressions that do not contain these new expression
forms. \emph{Values} \fbox{$f,g,A,B,F \in \D$} are expressions with no free de Bruijn
indices, where each neutral $n$ is under a reflection marker $\up A n$.
The types $A$ that direct reflection $\up A n$ and reification
$\down A f$ also live in the value world.
\[
\begin{array}{lllrl@{\qquad}l}
  \NeExp & \ni & n & ::= & \dots \mid \lev k
    & \mbox{de Bruijn level $k$} \\
  \tUp  & \ni & N & ::= & \up A n
     & \mbox{reflection of neutral term $n$ into the value world $\D$, at
       type $A$} \\
  \Whnf & \ni & w & ::= & \dots \mid N
    & \mbox{reflected neutral is weak head normal} \\
  \Exp  & \ni & t & ::= & \dots \mid \down A f
     & \mbox{reification of value $f$ at type $A$} \\
\end{array}
\]
% \[
% \begin{array}{lllrl@{\qquad}l}
% %  \Exp  & \ni & t & ::= & \dots \mid \lev k \mid \up T t \mid \down T t \\
%   \tUp  & \ni & N & ::= & \up A n
%      & \mbox{reflection of neutral term $n$ into the value world $\D$, at
%        type $A$} \\
%   \tDown & \ni & d & ::= & \down A f
%      & \mbox{reification of value $f$ at type $A$} \\
%   \NeExp & \ni & n & ::= & \dots \mid \lev k
%     & \mbox{de Bruijn level $k$} \\
%   \Whnf & \ni & w & ::= & \dots \mid N
%     & \mbox{reflected neutral are weak head normal} \\
%   \Exp  & \ni & t & ::= & \dots \mid d
%     & \mbox{reified values may occur in expressions} \\
% %  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
% \end{array}
% \]
\emph{De~Bruijn levels} are the mirror images of de Bruijn indices.  While de Bruijn
indices index the context from the right, \ie, $\ind 0$ refers the
last type that entered the context, de Bruijn levels index it from the
left, \ie, $\lev 0$ refers to the first type in the context.  This
way, de Bruijn levels are stable under context extensions, and
suitable to represent unknowns.

@@

Figure~\ref{fig:typedchart}, adapted from \citet{abel:habil} summarized the syntactic categories and main operations involved in NbE in what is called \emph{locally nameless style}.  The \textcolor{red}{red path} $\Exp \redto \D \redto \Dnf \redto \Nf$ decomposes $\beta\eta$-normalization into three steps.
\begin{enumerate}
\item
First, we close the term $t$ with an environment \textcolor{blue}{$\eta$} that maps the
free de Bruijn indices of $t$ to reflected de Bruijn levels.
Reflection of de Bruijn levels follows the \textcolor{blue}{blue path}
$\Level \blueto \DNe \blueto \D$: Levels embed via constructor $\var{}$
into semantic neutrals $\DNe$ which are labeled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.

@@

We define the operational semantics of our language by the weak head
evaluation relation \fbox{$t \evalsto w$} which is defined on
expressions, thus works on values as well as on terms.  It is defined mutually
with auxiliary relation \fbox{$w @ e \evalsto w'$} stating that weak
head normal form $w$ is eliminated by $e$ into weak head normal form $w'$.
% \[
% \begin{array}{l@{\qquad}l}
%   t \evalsto w & \mbox{term $t$ evaluates to weak head normal form $w$} \\
%   w @ e \evalsto w' & \mbox{eliminating $w$ with $e$ evaluates to $w'$} \\
% %  w @ \vec e \evalsto w' & \mbox{applying $w$ to eliminations $\vec e$ evaluates to $w'$} \\
% \end{array}
% \]

\fbox{$t \evalsto w$}
\vspace{-4ex}
\begin{gather*}
  \ru{}{w \evalsto w}
\qquad
  \ru{t \evalsto w \qquad w @ e \evalsto w'}{t\,e \evalsto w'}
\end{gather*}

% \fbox{$w @ \vec e \evalsto w'$}
% \begin{gather*}
%   \ru{}{w @ () \evalsto w}
% \qquad
%   \ru{w @ e \evalsto w' \qquad
%       w' @ \vec e \evalsto w''
%     }{w @ (e, \vec e) \evalsto w''}
% \end{gather*}

\fbox{$w @ e \evalsto w'$}
\begin{gather*}
  \ru{t[u] \evalsto w
    }{(\lambda t) @ u \evalsto w}
\qquad
  \ru{t[\alpha] \evalsto w
    }{(\lambda t) @ \alpha \evalsto w}
\qquad
  \ru{t[\alpha] \evalsto w
    }{(\lambda t) @ \ann \alpha \evalsto w}
\qquad
  \ru{t_z \evalsto w
    }{(\zero \alpha) @ \casett \evalsto w}
\\[2ex]
  \ru{t_s\, t \evalsto w
    }{(\suc \alpha t) @ \casett \evalsto w}
\qquad
  \ru{t\,\ann \alpha\,(\lambda x.\, x\, \fixtt)\,c \evalsto w
     }{c @ \fixtt \evalsto w
     }{c\in \{ \zero \alpha, \suc \alpha u \}}
\end{gather*}
%The rules added for NbE deal with elimination of delayed reflection:
For NbE, we add evaluation rules that deal with elimination of delayed reflection:
\begin{gather*}
  \ru{A' \evalsto \piT A B
    }{(\up {A'} n) @ u \evalsto \up{B[u]} (n \, \down A u) }
\qquad
  \ru{A \evalsto \piT \Size B
    }{(\up {A} n) @ \alpha \evalsto \up{B[\alpha]} (n \, \alpha)}
\qquad
  \ru{A \evalsto \forallT B
    }{(\up {A} n) @ \ann \alpha \evalsto \up{B[\alpha]} (n \ann \alpha)}
\\[2ex]
  \ru{%A \evalsto \Nat \beta \qquad
    }{(\up {A} n) @ \case[\ell] B {f_z} {f_s} \evalsto
      \up{B\,(\up{A} n)}\,
      % \up{B\,(\up{\Nat \infty} n)}
         n\,\case[\ell]
             {(\down {\Nat \infty \to \Set[\ell]} B)}
             {(\down {B\,{\zero\infty}} f_z)}
             {(\down {\funS x {\Nat \infty} B\,(\suc \infty x)} f_s)}
    }
\\[2ex]
  \ru{%A \evalsto \Nat \beta \qquad
    }{(\up {A} n) @ \fix[\ell] B f \evalsto
      n\,\fix[\ell] {(\down {\FixK} B)} {(\down {\FixT\, B} f)}
    }
\end{gather*}

@@

\noindent
\fbox{$\RSize k {\alpha} a$}\ \ Read back size value $\alpha$.
\begin{gather*}
  \ru{
    }{\RSize k \infty \infty}
\qquad
  \ru{
    }{\RSize k o o}
\qquad
  \ru{
    }{\RSize k {\lev j + o} {\ind{k\dotminus(1+j)} + o}}
\end{gather*}

@@

%The PERs $\NE$ characterizes neutral reifiable terms.
% \[
% \begin{array}{lll}%{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
%   \fbox{$\up T n = \up {T'} {n'} \in \NE$}
%     & \defas &
%     T \evalsto N \mand T' \evalsto N' \mand
%     \mforall k \mboth
%     \RNe k n m \mand \RNe k {n'} {m'} \mand m \nfeq m'
%     .\\
% \end{array}
% \]
The PERs $\NE$ and $\NF$ characterize (neutral) normalizing terms.
\[
% \begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
%   n = n' \in \NE
%     & \defas &
%     \RNe k n m &\mand& \RNe k {n'} {m} &\mforall k \\
%   d = d' \in \NF
%     & \defas &
%     \R k d v &\mand& \R k {d'} {v} &\mforall k \\
%   e = e' \in \ELIM
%     & \defas &
%     \RElim k e e_v &\mand& \R k {e'} {e_v} &\mforall k \\
%   A = A' \in \TY
%     & \defas &
%     \RTy k A V &\mand& \RTy k {A'} {V} &\mforall k \\
% \end{array}
%\newcommand{\myfbox}[1]{\begin{array}[c]{@{}l@{}}\framebox(60,12)[l]{~\,#1}\end{array}}
\newcommand{\myfbox}[1]{\framebox(60,12)[l]{~\,#1}}
\begin{array}{l@{\quad}l@{\quad}l@{~}l@{\quad}l@{~}l@{~\;}r@{~}l@{~}l@{~}l}
  \raisebox{-0.5ex}{\myfbox{$n = n' \in \NE$}}
    & \defas &
    \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m &\nfeq& m' &\mforall k
\\[1ex]
  \raisebox{-0.8ex}{\myfbox{$d = d' \in \NF$}}
    & \defas &
    \R k d v &\mand& \R k {d'} {v'} &\mand& v &\nfeq& v' &\mforall k
\\[1ex]
  \raisebox{-0.6ex}{\myfbox{$e = e' \in \ELIM$}}
    & \defas &
    \RElim k e e_v &\mand& \R k {e'} {e_v'} &\mand& e_v &\nfeq& e_v' &\mforall k
\\[1ex]
  \raisebox{-0.8ex}{\myfbox{$A = A' \in \TY$}}
    & \defas &
    \RTy k A V &\mand& \RTy k {A'} {V'} &\mand& V &\nfeq& V' &\mforall k \\
\end{array}
  % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \nfeq m' \mforall k
  %   }{t = t' \in \NE}
  % \ru{t \evalsto \up N n \qquad t' \evalsto \up{N'}{n'} \qquad n \nfeq n'
  %   }{t = t' \in \NE}
\]
%All PERs $\A$ we consider are sandwiched $\NE \subseteq \A \subseteq \NF$ between $\NE$ and $\NF$.

@@

By induction on $\ell \in \NN$ we define the PER family
$\_ = \_ \in \Set[\ell]$ of types
together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
which is a PER of values of type $A$.
The rules for \fbox{$T = T' \in \SET[\ell]$} are:
\[
\begin{array}{l@{\qquad}l}
  \ru{% T \evalsto \up{\Set[\ell']} n \qquad
      % T' \evalsto \up{\Set[\ell'']} n' \qquad
      % n = n' \in \NE
      T = T' \in \NEE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NEE
\\[3ex]
  \ru{T \evalsto \Nat\,\alpha \qquad T' \evalsto \Nat\,\alpha
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NAT(\alpha)
\\[3ex]
  \rux{T \evalsto \Set[\ell'] \qquad T' \evalsto \Set[\ell']
     }{T = T' \in \SET[\ell]
     }{\ell'{<}\ell}
  & \EL[\ell](T) = \SET[\ell']
\\[3ex]
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \\
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\EL[\ell](A),\;u \mapsto \EL[\ell](B[u]))
\\[3ex]
  \rul{T \evalsto \piT \Size B \quad
      T' \evalsto \piT{\Size}{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\SIZE,\; \alpha \mapsto \EL[\ell](B[\alpha]))
\\[3ex]
  \rul{T \evalsto \forallT B \quad
      T' \evalsto \forallT{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \FORALL(\alpha \mapsto \EL[\ell](B[\alpha]))
\end{array}
\]
All relations involved here are closed under weak head equality.
\begin{lemma}[Well-definedness]
  Let $\DD :: T_1 = T_2 \in \SET[\ell]$.
  \begin{enumerate}
  \item Symmetry: $T_2 = T_1 \in \SET[\ell]$.
  \item Transitivity: If\/ $T_2 = T_3 \in \SET[\ell]$ then $T_1 = T_3 \in \SET[\ell]$.
  \item Extension: $\EL[\ell](T_1) = \EL[\ell](T_2)$ and ``both'' are PERs.
  \end{enumerate}
\end{lemma}
\LONGVERSION{
\begin{proof}
  Simultaneously by induction on $\DD$.
\end{proof}
}

% The PER model has the usual properties, one of them being that $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$.  Thus, we simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.
\begin{lemma}[Derivation independence of extension]
\label{lem:indep}
  If\/ $\DD_1 :: T = T_1 \in \SET[\ell_1]$
  and $\DD_2 :: T_2 = T \in \SET[\ell_2]$ then $\EL[\ell_1](T) = \EL[\ell_2](T)$.
\end{lemma}
\LONGVERSION{
\begin{proof}
  By induction on $\DD_1$ and cases on $\DD_2$.
\end{proof}
}
Since $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$,  we may simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.

@@

Reflection and reification perform $\eta$-expansion such that we arrive at an $\eta$-long $\beta$-normal form.  To perform the $\eta$-expansion, not the precise type is needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
For the logical framework, the shape of a dependent type is just its
underlying simple type \cite{harperPfenning:equivalenceLF}. However,
in the presence of universes and large eliminations, there is no
underlying simple type.  Of course, we can take a type as its own
shape, but we want at least that $\Nat \alpha$ and $\Nat \beta$ have
the same shape even for different $\alpha, \beta$.
Also all neutral types can be summarized under a single shape.

@@

However, templates are not closed under subtyping in either direction because of contravariance.

@@

%A type constructor $B \in \FixK\,\ell$ is \emph{admissible for recursion}, \fbox{$B \in \ADM$}, if
%$\bigcap_{
Let us write \fbox{$B = B' \in \Adm\ell$} iff $B = B' \in \FixK\,\ell$ and
for all $\beta \in \SIZE$ and $a \in \Nat \beta$ we have
$B\,\beta\, a \leq B\,\infty\,a \in \TYPE$ and
$B'\,\beta\, a \leq B'\,\infty\,a \in \TYPE$.
\begin{lemma}[Fix]
Let $g = a \, \fix[\ell] B f$ and $g' = a'\,\fix[\ell]{B'}{f'}$.
If\/ $a = a' \in \Nat \alpha$
and $B = B' \in \Adm\ell$
and $f = f' \in \FixT\,B$
then $g = g' \in B\,\alpha\,a$.
\end{lemma}
\begin{proof*}
By well-founded induction on $\alpha$.
\begin{caselist}

@@

%% Bibliography
\bibliography{auto-icfp17}

>>>>

In this section we present an operational semantics of our language,
define the NbE algorithm, construct a PER model, and demonstrate that
NbE is complete for definitional equality, \ie, if
$\Gamma \der t = t' : T$, then $t$ and $t'$ have the same normal form
up to annotations.
\[
\begin{array}{lllll@{\qquad}l}
  \Ne & \ni &
  m & ::= & \ind i \mid m\,v \mid m\,a \mid m\,\ann a \mid
            m\,\case[\ell] V{v_z}{v_s} \mid m\,\fix[\ell] V v
    & \mbox{normal neutral term} \\
  \Nf & \ni &
  v & ::= & m \mid \lambda v \mid \zero a \mid \suc a v \mid
            \Set[\ell] \mid \Nat a \mid \piT {V_u}{V_t} \mid
            \epiT\Size V
    & \mbox{normal form} \\
\end{array}
\]
For the operational semantics,
instead of defining a separate language of values,
we extend the syntax of expressions by de Bruijn levels $\lev k$
to be used as generic values (unknowns), and type annotations
$\up A n$ and $\down A t$ for lazy realizations of the
reflection and reification operations of NbE.
\emph{Terms} are expressions that do not contain these new expression
forms. \emph{Values} \fbox{$f,g,A,B,F \in \D$} are expressions with no free de Bruijn
indices, where each neutral $n$ is under a reflection marker $\up A n$.
The types $A$ that direct reflection $\up A n$ and reification
$\down A f$ also live in the value world.
\[
\begin{array}{lllrl@{\qquad}l}
  \NeExp & \ni & n & ::= & \dots \mid \lev k
    & \mbox{de Bruijn level $k$} \\
  \tUp  & \ni & N & ::= & \up A n
     & \mbox{reflection of neutral term $n$ into the value world $\D$, at
       type $A$} \\
  \Whnf & \ni & w & ::= & \dots \mid N
    & \mbox{reflected neutral is weak head normal} \\
  \Exp  & \ni & t & ::= & \dots \mid \down A f
     & \mbox{reification of value $f$ at type $A$} \\
\end{array}
\]
% \[
% \begin{array}{lllrl@{\qquad}l}
% %  \Exp  & \ni & t & ::= & \dots \mid \lev k \mid \up T t \mid \down T t \\
%   \tUp  & \ni & N & ::= & \up A n
%      & \mbox{reflection of neutral term $n$ into the value world $\D$, at
%        type $A$} \\
%   \tDown & \ni & d & ::= & \down A f
%      & \mbox{reification of value $f$ at type $A$} \\
%   \NeExp & \ni & n & ::= & \dots \mid \lev k
%     & \mbox{de Bruijn level $k$} \\
%   \Whnf & \ni & w & ::= & \dots \mid N
%     & \mbox{reflected neutral are weak head normal} \\
%   \Exp  & \ni & t & ::= & \dots \mid d
%     & \mbox{reified values may occur in expressions} \\
% %  \Whnf & \ni & w,d,f,g,A,B,F,G & ::= & \dots \mid  \up T n \mid \down T t \\
% \end{array}
% \]
\emph{De~Bruijn levels} are the mirror images of de Bruijn indices.  While de Bruijn
indices index the context from the right, \ie, $\ind 0$ refers the
last type that entered the context, de Bruijn levels index it from the
left, \ie, $\lev 0$ refers to the first type in the context.  This
way, de Bruijn levels are stable under context extensions, and
suitable to represent unknowns.

@@

Figure~\ref{fig:typedchart}, adapted from \citet{abel:habil} summarized the syntactic categories and main operations involved in NbE in what is called \emph{locally nameless style}.  The \textcolor{red}{red path} $\Exp \redto \D \redto \Dnf \redto \Nf$ decomposes $\beta\eta$-normalization into three steps.
\begin{enumerate}
\item
First, we close the term $t$ with an environment \textcolor{blue}{$\eta$} that maps the
free de Bruijn indices of $t$ to reflected de Bruijn levels.
Reflection of de Bruijn levels follows the \textcolor{blue}{blue path}
$\Level \blueto \DNe \blueto \D$: Levels embed via constructor $\var{}$
into semantic neutrals $\DNe$ which are labeled with their type $A \in \D$
to become an element $\up A \lev j \in \D$.

@@

We define the operational semantics of our language by the weak head
evaluation relation \fbox{$t \evalsto w$} which is defined on
expressions, thus works on values as well as on terms.  It is defined mutually
with auxiliary relation \fbox{$w \at e \evalsto w'$} stating that weak
head normal form $w$ is eliminated by $e$ into weak head normal form $w'$.
% \[
% \begin{array}{l@{\qquad}l}
%   t \evalsto w & \mbox{term $t$ evaluates to weak head normal form $w$} \\
%   w \at e \evalsto w' & \mbox{eliminating $w$ with $e$ evaluates to $w'$} \\
% %  w \at \vec e \evalsto w' & \mbox{applying $w$ to eliminations $\vec e$ evaluates to $w'$} \\
% \end{array}
% \]

\noindent
\fbox{$t \evalsto w$}
\ and \
\fbox{$w \at e \evalsto w'$}
\vspace{-6ex}
\begin{gather*}
  \ru{}{w \evalsto w}
\qquad
  \ru{t \evalsto w \qquad w \at e \evalsto w'}{t\,e \evalsto w'}
% \end{gather*}
% \fbox{$w \at e \evalsto w'$}
% \begin{gather*}
\\[1.5ex]
  \ru{t[u] \evalsto w
    }{(\lambda t) \at u \evalsto w}
\qquad
  \ru{t[\alpha] \evalsto w
    }{(\lambda t) \at \alpha \evalsto w}
\qquad
  \ru{t[\alpha] \evalsto w
    }{(\lambda t) \at \ann \alpha \evalsto w}
\qquad
  \ru{t_z \evalsto w
    }{(\zero \alpha) \at \casett \evalsto w}
\\[1.5ex]
  \ru{t_s\, t \evalsto w
    }{(\suc \alpha t) \at \casett \evalsto w}
\qquad
  \rux{t\,\ann \alpha\,(\lambda x.\, x\, \fixtt)\,c \evalsto w
     }{c \at \fixtt \evalsto w
     }{c\in \{ \zero \alpha, \suc \alpha u \}}
\end{gather*}
%The rules added for NbE deal with elimination of delayed reflection:
For NbE, we add evaluation rules that deal with elimination of delayed reflection:
\begin{gather*}
  \ru{A' \evalsto \piT A B
    }{(\up {A'} n) \at u \evalsto \up{B[u]} (n \, \down A u) }
\qquad
  \ru{A \evalsto \piT \Size B
    }{(\up {A} n) \at \alpha \evalsto \up{B[\alpha]} (n \, \alpha)}
\qquad
  \ru{A \evalsto \forallT B
    }{(\up {A} n) \at \ann \alpha \evalsto \up{B[\alpha]} (n \ann \alpha)}
\\[1.2ex]
  \ru{%A \evalsto \Nat \beta \qquad
    }{(\up {A} n) \at \case[\ell] B {f_z} {f_s} \evalsto
      \up{B\,(\up{A} n)}\,
      % \up{B\,(\up{\Nat \infty} n)}
         n\,\case[\ell]
             {(\down {\Nat \infty \to \Set[\ell]} B)}
             {(\down {B\,{\zero\infty}} f_z)}
             {(\down {\funS x {\Nat \infty} B\,(\suc \infty x)} f_s)}
    }
\\[1ex]
  \ru{%A \evalsto \Nat \beta \qquad
    }{(\up {A} n) \at \fix[\ell] B f \evalsto
      n\,\fix[\ell] {(\down {\FixK} B)} {(\down {\FixT\, B} f)}
    }
\end{gather*}

@@

\noindent
\fbox{$\RSize k {\alpha} a$}\ \ Read back size value $\alpha$.
\vspace{-1.5ex}
\begin{gather*}
  \ru{
    }{\RSize k \infty \infty}
\qquad
  \ru{
    }{\RSize k o o}
\qquad
  \ru{
    }{\RSize k {\lev j + o} {\ind{k\dotminus(1+j)} + o}}
\end{gather*}

@@

The PERs $\NE$ and $\NF$ characterize (neutral) normalizing values.
For instance, two values $n$ and $n'$ are related in $\NE$ if at any $k \in \NN$
they can be read back to neutral normal forms $m$ and $m'$
which are identical up to annotations.
\[
% \begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
%   n = n' \in \NE
%     & \defas &
%     \RNe k n m &\mand& \RNe k {n'} {m} &\mforall k \\
%   d = d' \in \NF
%     & \defas &
%     \R k d v &\mand& \R k {d'} {v} &\mforall k \\
%   e = e' \in \ELIM
%     & \defas &
%     \RElim k e e_v &\mand& \RElim k {e'} {e_v} &\mforall k \\
%   A = A' \in \TY
%     & \defas &
%     \RTy k A V &\mand& \RTy k {A'} {V} &\mforall k \\
% \end{array}
%\newcommand{\myfbox}[1]{\begin{array}[c]{@{}l@{}}\framebox(60,12)[l]{~\,#1}\end{array}}
\newcommand{\myfbox}[1]{\framebox(60,13)[l]{~\,#1}}
\begin{array}{l@{\quad}l@{\quad}l@{~}l@{\,}l@{\quad}l@{\,}l@{~}l@{~\;}r@{~}l@{~}l@{~}l}
  \raisebox{-0.5ex}{\myfbox{$n = n' \in \NE$}}
    & \defas &
    \tRne _k & n \evalsto m &\mand& \tRne _k & {n'} \evalsto {m'} &\mand& m &\nfeq& m' &\mforall k
\\[1ex]
  \raisebox{-0.8ex}{\myfbox{$d = d' \in \NF$}}
    & \defas &
    \tR _k & d \evalsto v &\mand& \tR _k & {d'} \evalsto {v'} &\mand& v &\nfeq& v' &\mforall k
\\[1ex]
  \raisebox{-0.6ex}{\myfbox{$e = e' \in \ELIM$}}
    & \defas &
    \tRelim _k & e \evalsto e_v &\mand& \tRelim _k & {e'} \evalsto {e_v'} &\mand& e_v &\nfeq& e_v' &\mforall k
\\[1ex]
  \raisebox{-0.8ex}{\myfbox{$A = A' \in \TY$}}
    & \defas &
    \tRty _k & A \evalsto V &\mand& \tRty _k & {A'} \evalsto {V'} &\mand& V &\nfeq& V' &\mforall k \\
\end{array}
% \begin{array}{l@{\quad}l@{\quad}l@{~}l@{\quad}l@{~}l@{~\;}r@{~}l@{~}l@{~}l}
%   \raisebox{-0.5ex}{\myfbox{$n = n' \in \NE$}}
%     & \defas &
%     \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m &\nfeq& m' &\mforall k
% \\[1ex]
%   \raisebox{-0.8ex}{\myfbox{$d = d' \in \NF$}}
%     & \defas &
%     \R k d v &\mand& \R k {d'} {v'} &\mand& v &\nfeq& v' &\mforall k
% \\[1ex]
%   \raisebox{-0.6ex}{\myfbox{$e = e' \in \ELIM$}}
%     & \defas &
%     \RElim k e e_v &\mand& \RElim k {e'} {e_v'} &\mand& e_v &\nfeq& e_v' &\mforall k
% \\[1ex]
%   \raisebox{-0.8ex}{\myfbox{$A = A' \in \TY$}}
%     & \defas &
%     \RTy k A V &\mand& \RTy k {A'} {V'} &\mand& V &\nfeq& V' &\mforall k \\
% \end{array}
\]
Once we have established useful closure properties of these PERs, they
abstract most of the reasoning about the read-back relation from our
proofs.  This idea is due to Coquand \cite{abelCoquandPagano:tlca09}.

@@

By induction on $\ell \in \NN$ we define the PER family
$\_ = \_ \in \Set[\ell]$ of types
together with the extension $\EL[\ell] T$ (for $T = T' \in \Set[\ell]$)
which is a PER of values of type $T$.
The rules for \fbox{$T = T' \in \SET[\ell]$} are:
\[
\begin{array}{l@{\qquad}l}
  \ru{% T \evalsto \up{\Set[\ell']} n \qquad
      % T' \evalsto \up{\Set[\ell'']} n' \qquad
      % n = n' \in \NE
      T = T' \in \NEE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NEE
\\[3ex]
  \ru{T \evalsto \Nat\,\alpha \qquad T' \evalsto \Nat\,\alpha
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \NAT(\alpha)
\\[3ex]
  \rux{T \evalsto \Set[\ell'] \qquad T' \evalsto \Set[\ell']
     }{T = T' \in \SET[\ell]
     }{\ell'{<}\ell}
  & \EL[\ell](T) = \SET[\ell']
\\[3ex]
  \rul{T \evalsto \piT A B \qquad
      T' \evalsto \piT{A'}{B'} \\
      A = A' \in \SET[\ell] \qquad
      B[u] = B'[u'] \in \SET[\ell] \mforall u = u' \in \EL[\ell](A)
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\EL[\ell](A),\;u \mapsto \EL[\ell](B[u]))
\\[3ex]
  \rul{T \evalsto \piT \Size B \quad
      T' \evalsto \piT{\Size}{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \PI(\SIZE,\; \alpha \mapsto \EL[\ell](B[\alpha]))
\\[3ex]
  \rul{T \evalsto \forallT B \quad
      T' \evalsto \forallT{B'} \quad
      B[\alpha] = B'[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
    }{T = T' \in \SET[\ell]}
  & \EL[\ell](T) = \FORALL(\alpha \mapsto \EL[\ell](B[\alpha]))
\end{array}
\]
All relations involved here are closed under weak head equality.
\begin{lemma}[Well-definedness]
  Let $\DD :: T_1 = T_2 \in \SET[\ell]$.
  \begin{enumerate}
  \item Symmetry: $T_2 = T_1 \in \SET[\ell]$.
  \item Transitivity: If\/ $T_2 = T_3 \in \SET[\ell]$ then $T_1 = T_3 \in \SET[\ell]$.
  \item Extension: $\EL[\ell](T_1) = \EL[\ell](T_2)$ and ``both'' are PERs.
  \end{enumerate}
\end{lemma}
\LONGVERSION{
\begin{proof}
  Simultaneously by induction on $\DD$.
\end{proof}
}

% The PER model has the usual properties, one of them being that $\EL[\ell](T)$ does not depend $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$.  Thus, we simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.
\begin{lemma}[Derivation independence of extension]
\label{lem:indep}
  If\/ $\DD_1 :: T = T_1 \in \SET[\ell_1]$
  and $\DD_2 :: T_2 = T \in \SET[\ell_2]$ then $\EL[\ell_1](T) = \EL[\ell_2](T)$.
\end{lemma}
\LONGVERSION{
\begin{proof}
  By induction on $\DD_1$ and cases on $\DD_2$.
\end{proof}
}
Since $\EL[\ell](T)$ does not depend on $\ell$ nor the derivation that introduced $T = T' \in \SET[\ell]$,  we may simply write $t = t' \in \EL(T)$ or even $t = t' \in T$.

@@

Reflection and reification perform $\eta$-expansion so that we arrive at an $\eta$-long $\beta$-normal form.  To perform the $\eta$-expansion, not the precise type is needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
For the logical framework, the shape of a dependent type is just its
underlying simple type \cite{harperPfenning:equivalenceLF}. However,
in the presence of universes and large eliminations, there is no
underlying simple type.  Of course, we can take a type as its own
shape, but we want at least that $\Nat \alpha$ and $\Nat \beta$ have
the same shape even for different $\alpha, \beta$.
Also all neutral types can be summarized under a single shape.

@@

However, templates are not closed under subtyping in either direction because subtyping is contravariant for function type domains but the shape relation is covariant.

@@

%A type constructor $B \in \FixK\,\ell$ is \emph{admissible for recursion}, \fbox{$B \in \ADM$}, if
%$\bigcap_{
\noindent
Let us write \fbox{$B = B' \in \Adm\ell$} iff $B = B' \in \FixK\,\ell$ and
for all $\beta \in \SIZE$ and $a \in \Nat \beta$ we have
$B\,\beta\, a \leq B\,\infty\,a \in \TYPE$ and
$B'\,\beta\, a \leq B'\,\infty\,a \in \TYPE$.
\begin{lemma}[Fix]
Let $g = a \, \fix[\ell] B f$ and $g' = a'\,\fix[\ell]{B'}{f'}$.
If\/ $a = a' \in \Nat \alpha$
and $B = B' \in \Adm\ell$
and $f = f' \in \FixT\,B$
then $g = g' \in B\,\alpha\,a$.
\end{lemma}
\begin{proof*}
By well-founded induction on $\alpha$.
\begin{caselist}

@@

%% Bibliography
\bibliography{auto-icfp17-long}
